Concurrent Programming with CXL

### 2020-05-20 book/paper.tex
Better explanation of TAS

\begin{figure}
\begin{code}
\verbatiminput{UpLock.cxl}
\end{code}
\caption{Na\"{\i}ve implementation of a lock.}
\label{fig:uplock}
\end{figure}

@@

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process $i$ is in the critical
section, \texttt{turn} == $i$.  Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{lk} points to a shared variable holding a
boolean.  The operation assigns \texttt{True} to the
variable but returns the value the variable had \emph{before}
the assignment.
Figure~\ref{fig:tas} goes on to show how the program
in Figure~\ref{fig:uplock} can be made correct.
Using \texttt{tas} a process can wait until \texttt{lock} is
\texttt{False} and then set it to \texttt{True} before another
process is able to see its old value.

Compared to Peterson's Algorithm, this approach is general for
any number of processes.  To make sure, you may want to try it
for three or more processes.  In fact, you may want to replace
the last two lines by:
\begin{code}
\begin{verbatim}
for i in 1..5:
    spawn process(), ();
;
\end{verbatim}
\end{code}
This creates 5 processes with the same name tag.  CXL can
exploit their symmetry to do a much more efficient state
search.  Try replacing it with the following code and see
the difference:
\begin{code}
\begin{verbatim}
for i in 1..5:
    spawn process(), i;
;
\end{verbatim}
\end{code}

>>>>

\begin{figure}
\begin{code}
\verbatiminput{UpLock.cxl}
\end{code}
\caption{Na\"{\i}ve implementation of a shared lock.}
\label{fig:uplock}
\end{figure}

@@

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process $i$ is in the critical
section, \texttt{turn} == $i$.
(If you do not believe that, replace the assertion at the critical
section with \texttt{assert turn == self} and run CXL.)
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable and then sets the shared variable to \texttt{True}.

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of the variables is \texttt{False}.
Another invariant, $I_2$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two, it is clear that only one process can be in the
critical section at the same time.

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.

\begin{figure}
\begin{code}
\verbatiminput{../TAS.cxl}
\end{code}
\caption{Fixed version of Figure~\ref{fig:uplock} using test-and-set.}
\label{fig:tas2}
\end{figure}

Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False} and no other process modified
\texttt{private[$i$]}.
Indeed, there is no need to keep \texttt{private} as a shared
variable---we only did so to be able to show the invariants.
Figure~\ref{fig:tas2} shows a more straightforward implementation.
Compared to Peterson's Algorithm, this approach is general for
any number of processes.
